---
title: "HW 5 solutions"
author: "Drew Steen"
date: "9/15/2020"
output: md_document
---

# Setup
```{r}
library(tidyverse)
library(nycflights13)
```


# Split-apply-combine

1. Use `?nycflights13::flights` to see the documentation of flights. (You can't really show this in an RMarkdown document, but please do it.)

2. Use the `glimpse()` function to get a general sense of the `flights` data frame. How many rows and columns does `flights` have? 

```{r}
glimpse(flights)
```

We can see from `glimpse()` that `flights` has `r nrow(flights)` rows (observations). 

3. Use `filter()` to create a data frame of just the flights that departed 2 hours late or more. What fraction of flights departed at least 2 hours late?
```{r}
delays <- flights %>%
  filter(dep_delay > 120)
frac.late <- nrow(delays) / nrow(flights)
frac.late
```

We can see that `r formatC(frac.late * 100, type = "f", digits = 0)`% of flights were late.

4. Use `mutate()` and `paste0()` to create a column named `flight.code` that contains the combined airline and flight number (e.g., "UA5041").

```{r}
flights %>%
  mutate(flight.code = paste0(carrier, flight))
```

5. Use `group_by()`, `summarise()`, and `n()` to make a data frame showing how many flights each aircraft (`tail_num()`) flew for each carrier throughout the whole data set.

```{r}
frequent_fliers <- flights %>%
  group_by(carrier, tailnum) %>%
  summarise(n.flights = n())
  
```




