---
title: 'Lists'
author: ~
date: '2020-10-12'
slug: lists
categories: [base R]
tags: [week08]
draft: no
---

Working R's most flexible data type: lists.

<!--more-->

We have done a lot with vectors, and with data frames (including their tidyverse manifestation, tibbles). Now we're going to talk about the most flexible data type, lists.

(Side note: As [Groremund and Wickham point out](https://r4ds.had.co.nz/vectors.html#vector-basics), basically everything in R is a list, including atomic vectors. This is mostly a pedantic point, and for the most part it makes sense to think of a list as a separate thing from vectors). 

If vectors are like a 1-dimensional column of numbers, and data frames are like 2-dimensional data tables, I think of lists as being like a big bag: you can put whatever you want into them, and the shape is not important. However, lists are ordered - just like vectors (which, again, are a specific type of list!) - so there is a first element, a second element, etc.

Lists are great when you need to store poorly-structured data. One example you will probably run into a lot is the linear model object, which is a specific type of list. We can create a linear model (a.k.a. a linear regression) using the `lm()` function. For instance, say we want to make a linear model of care mileage, in miles per gallon, versus weight:

```{r}
my_mod <- lm(mpg ~ wt, data = mtcars)
str(my_mod)
```
Good golly, that's a lot of information in there. A statistician would say that a linear model is fully specified by just a slope and an intercept. But this contains all kinds of other information!

# Accessing list elements

There are three ways to access elements of a list:

* using `[`
* using `[[`
* using `$`

## Acessing list elements with `[`

Vectors are a type of list, so we can subset them like we subset lists:

```{r}
my_mod[1]
```

This looks useful - it has the word "coefficients" in there - but let's use `str()` to learn more.

```{r}
str(my_mod[1])
```

That's a little odd: it is a list of length one. It turns out that `[` subsetting lists *always returns another list* which contains the list element. This is usually not what we want. Instead...

## Subsetting lists with `[[`

...we can subset lists using `[[`, to get the actual contents of the respective list element. For instance,

```{r}
my_mod[[1]]
```

`my_mod[[1]]` returns a 2-element, named vector (which contains the slope and intercept of the linear model), while `my_mod[1]` returns *a list which contains* that same 2-element, named vector. 

**When you are accessing list elements, you will almost always want to use `[[` rather than `[` to access specific elements.**

## Accessing list elements by name

List elements are always numbered. They *can* also be named, although this is not required. For instance, the first element of an `lm` object is always named `coefficients`:

```{r}
my_mod[["coefficients"]]
```

`my_mod[["coefficients"]]` is exactly equivalent to `my_mod[[1]]`. 

We can see a bit more about this by creating our own lists using the `list()` function:

```{r}
my_list <- list(a_data_frame = mtcars, a.vector = letters, my_mod)
```

Here I've created a list of length 3 (try `length(my_list)`), where I named the first two elements `a_data_frame` and `a.vector` (check the style guide to remember why I used `_` in the data frame name and `.` in the vector name). But I didn't name the third element. So I can access the first two elements by name (`my_list[["a.vector"]]`) but I have to access the third element by index (`my_list[[3]]`). 

## Nested lists

Remember that lists can contain anything, including other lists. That means that list elements may be nested. This is what we saw in the `lm` object list - `str` showed us each element in each element of the list. An `lm` object goes 2 levels deep, but lists can in principle go as deep as you like.

You can access these deeper levels of lists by stacking subsetting notation:

```{r}
# This returns the actual data in the `mpg` vector from which the model was created.
my_mod[["model"]][["mpg"]]
```


# Acessing list elements with `$`

Finally, we can use `$` to access list elements by name:

```{r}
my_mod$coefficients
```

Note that when I used `[[` to access list elements by name, I had to surround the list elements with quotation marks, but when I use `$` syntax, I don't use quotation marks. At this point it is not worth thinking too hard about why that is, but you should keep it in mind.

## Accessing data frame columns with `$`

We come back to the point that data frames are a special kind of list, where each column-vector in the data frame is an element of the list. 

Since we can access list elements by name using `$`, we can access columns of data frames the same way. You will find that this is the most common way to access columns of data frames. For instance,

```{r}
str(mtcars$mpg)
```

shows that `mtcars$mpg` returns a vector corresponding to the `mpg` column of `mtcars`. 

We have also subsetted columns of data frames using the tidyverse function `select()`. There is an important difference though: `select()` always returns a **data frame**, whereas `$` and `[[` return **vectors**. 