---
title: "Data_frames"
date: 2020-09-10T16:45:24-04:00
draft: true
---

# This is my new test page

Discusses the data structure called data frames: what they are and why they are so important.

<!--more-->

# This is my new test page

```{r setup, echo=TRUE, message=FALSE}
library(tidyverse)
knitr::opts_chunk$set(eval=FALSE)
```

# Review of data structures

In lecture on Wednesday, Sept 9, we discussed a key data structure in R: **vectors**^[Properly speaking, these are called atomic vectors. However, colloquially we refer to these as vectors.]. There are three important rules for vectors:

* They are **1-dimensional**: think of them like a column of values.
* They are **ordered**: there is a first element, a second element, etc.
* They contain **a single data type**: a vector can be integer, double, logical, or character, but they can't be a mix of those. If a vector looks like it has a mix of data types, that's because it has been *coerced* to a more general type. For instance, `c(7, "aardvark")` will be coerced to a character vector: the number `7` will be converted to the character `"7"`.

There are a two other data structures which are frequently used in base R, but not in the tidyverse:

* **Matrix**: a 2-dimensional vector. This follows the same rules as vectors, but it has rows and columns.
* **List**: I think of this like a bag, into which you can throw any data structure you want. (Lists are ordered, like vectors, so maybe it is an unusually well-ordered bag?) The first element of a list might be a character vector, the second might be a integer matrix, and the third might be another list (which might, in turn, contain more lists)! These are great when you have to keep track of non-parallel information about something. For instance, a linear model object contains the coefficients of the linear model,  the residuals, and all sorts of other information that comes in different data types and different lengths. The `lm()` function therefore returns a special kind of list to store all that information.

# Data frames and tibbles

A data frame is a special kind of list. Data frames must obey two rules:

* The list can only contain **vectors**
* Each vector must be the same **length**

Therefore, data frames act like a rectangular table. For instance, here are  the first six rows of the data frame `mtcars`, a table that is built into R's **datasets** package, which is autoloaded when you start R:

```{r, echo = TRUE}
print(mtcars) # print the first 6 rows
```

```{r}
Sys.sleep(60)
```


For instance, say we were foresters, taking data about trees. Perhaps we would record the species of the tree, its height, whether it hosts parasites, and what we think its name would be if it were an [Ent](https://en.wikipedia.org/wiki/Ent). 

Each quantity we measure would be recorded in a separate column of the data frame, and each 
